A correct by construction hierarchy of reflective tactics for proving equalities in algebraic structures

### 2014-07-03 newPaper/content/introduction.tex
paper...

Proving that one term is equal to another within some theories is something quite common in formal certification. Of course, the two things are usually not syntactically equal --this case would be trivial--, but they are equal according to a set of properties. These properties are either obtained by implementation (the implementation of + for relatives numbers is symmetric for example), or by axioms if we are working with an abstract axiomatised structure. 
These proofs are usually omitted "on the paper" for the ordinary every day maths, but they are required by proof assistants, no matter how obvious they are. Currently, in the Idris\footnote{Idris is a relatively new, purely functional, and dependently typed programming language with full dependent types, which also supports tactic based proofs  (Idris will be briefly defined the the [Section X] (at the end of the paper))} programming language, they are done by hand, and they are made of a potentially long sequence of rewriting, where each step of rewriting uses one of the available properties. 
These proofs are the every day routine we are working in a dependently typed language, as Idris, and especially when we're intensively indexing types over values in order to capture certain logical properties.
The indices can be for example natural numbers, and the lemmas we often have to prove about these indices are proven by some simple --but potentially long-- rewritings which use the properties of natural number and the usual operations : symmetry of +, commutativity of +, etc.

\subsection{A motivating little example}
Let's define the types of bit and binary number, both indexed over the value they represent, expressed in base ten as a natural number.
\begin{code}[caption=Bit and binary number, captionpos=b, label=lst1:haskell2]
data Bit : Nat -> Type where
     b0 : Bit Z
     b1 : Bit (S Z)

@@

It will be useful to be able to pad a binary number expressed with $w$ bits to a binary number with $S\ w$ digits : 
\begin{code}[caption=Padding, captionpos=b, label=lst1:haskell2]
pad : Binary w n -> Binary (S w) n
pad zero = zero # b0
pad (num # x) = pad num # x
\end{code}
Of course, this operation preserves the value represented by the binary number : this function takes a $Binary\ w\ n$ and produces a $Binary\ (S\ w)\ n$, so there will be no need to prove the correctness of this function after its definition.

Now, we're tempted to write the addition between two binary numbers.
We start with an auxiliary function which adds three bits (the third one will play the role of the carry), and produces the two bits of the result.

@@

Again, we note that there is no need to produce any lemma of correctness about $addBit$ afterwards : the correct by construction style in which it is written already gives the property that we want : that two bits produced effectively represent the addition of the three bits given in input.

Now, we would like to define a function adding two binary numbers and a bit (used as a carry). We demand the two binary number to be expressed with the same width $w$, and we produce a result of the same width, which of course should represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{code}[caption=Addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{code}
Unfortunately, the types are mismatching for both of the two lines. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type $Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)$ and we're providing a term of type $Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))$ [I AM SURPRISED HERE BY THE TYPE OF THE PROVIDED TERM].
The problem is even bigger for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but we will have to prove that the term provided can be converted into a term which has the right type.
However, for both lines, it is possible to prove the equality between the expected and the provided types, so it is possible to convert the type of the provided term into the expected term. These proofs will be made of a series of rewriting, and each rewriting will use a property about natural numbers and the usual operations.
FOr example, the proof for the second line look like this :
\begin{code}[caption=Proof by hand of correspondence between expected and provided types for the second line of the definition of adc, captionpos=b, label=lst1:haskell2]
Main.adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
}
\end{code}

@@

Algebraic structures like monoid, group, commutative group and ring have a very nice property : every element in them has a canonical represent. Thus these kind of proofs can be automatically produced by a tactic by simply computing the normal forms for each side of the equality, and then comparing them with the syntactical equality (Leibniz equality).

@@

\begin{code}[caption=Code we'd like to write for the addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry 
        by CommutativeGroupSolver
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb 
        by CommutativeGroupSolver
\end{code}
We would like to let the solver for commutative groups to find the proof on its own, and thereby, there will be no extra lemmas to prove afterwards.

\subsection{Our contribution}

@@

For our goal, we focus on some specific theories, which are the following algebraic structures : magmas, semi-group, monoid, group and Abelian group.
The user will just have to prove that his structure satisfies the properties of an algebraic structure, and he will be allowed to invoke the corresponding tactic.

>>>>

Proving that one term is equal to another within some theories is something quite common in formal certification. Of course, the two things are usually not syntactically equal --this case would be trivial--, but they are equal according to a set of properties. These properties are either obtained by implementation (the implementation of + for natural numbers satisfies the property of associativity for example), or by axioms if we are working within an axiomatised structure. 
These proofs are usually omitted "on the paper" for the ordinary every day maths, but they are required by proof assistants, no matter how obvious they are. Currently, in the Idris\footnote{Idris is a relatively new, purely functional, and dependently typed programming language with full dependent types, which also supports tactic based proofs  (Idris will be briefly defined the the [Section X] (at the end of the paper))} programming language, they are done by hand, and they are made of a potentially long sequence of rewriting, where each step of rewriting uses one of the available properties. 
These proofs are the every day routine when we are working in a dependently typed language, as Idris, and especially when we're intensively indexing types over values in order to capture certain logical properties.
The indices can be for example natural numbers, and in this case the lemmas that we have to prove about these indices are arithmetic statements which use the properties of natural numbers and their usual operations : symmetry and commutativity for +, existence of neutral element for +, etc.

\subsection{A motivating little example}
We introduce a little example which shows how these proofs appear when we index a type by values like natural numbers.
Let's define the types of $Bit$ and $Binary$ number, both indexed over the value they represent, expressed in base ten as a natural number.
\begin{code}[caption=Bit and binary number, captionpos=b, label=lst1:haskell2]
data Bit : Nat -> Type where
     b0 : Bit Z
     b1 : Bit (S Z)

@@

It will be useful to be able to pad a binary number expressed with $w$ bits to a binary number with $S\ w$ digits : 
\begin{code}[caption=Padding, captionpos=b, label=lst1:haskell2]
pad : Binary w n -> Binary (S w) n
pad zero = zero # b0
pad (num # x) = pad num # x
\end{code}
Of course, this operation preserves the value represented by the binary number : this function takes a $Binary\ w\ n$ and produces a $Binary\ (S\ w)\ n$, so there will be no need to prove the correctness of this function after its definition (the $value$ is still $n$).

Now, we're tempted to write the addition between two binary numbers.
We start with an auxiliary function which adds three bits (the third one will latter play the role of the carry), and produces the two bits of the result.

@@

Again, we note that there is no need to produce any lemma of correctness about $addBit$ afterwards : the correct by construction style in which it is written already gives the property that we want : that the two bits produced effectively represent the addition of the three bits given in input.

Now, we want to define a function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and we produce a result with one more bit, and this result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{code}[caption=Addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{code}
Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] and we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\] [I AM SURPRISED HERE BY THE TYPE OF THE PROVIDED TERM].
The problem is even worse for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition --provisional definitions are described with the brief overview of the language in page XX-- with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but we will have to prove afterwards that the term provided can be converted into a term which has the right type.
However, for both of these two patterns, it is possible to prove the equality between the expected and the provided types, and this is even stronger than we need. Using this equality of types it will be trivial to produce a term of the expected type by using the provided term and by simply rewriting its type with this proof. These proofs will be made of a series of rewriting, and each rewriting will use a property about natural numbers and the usual operations.
For example, the proof for the second pattern look like this :
\begin{code}[caption=Proof by hand of the correspondence between expected and provided types for the second line of the definition of adc, captionpos=b, label=lst1:haskell2]
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
}
\end{code}

@@

@@

\begin{code}[caption=Code we'd like to write for the addition of two binary numbers, captionpos=b, label=lst1:haskell2]
adc : Binary w x -> Binary w y -> Bit c 
     -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry 
        by CommutativeGroupSolver
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = 
      addBit bX bY carry in
          adc numx numy carry0 # lsb 
        by CommutativeGroupSolver
\end{code}
We would like to let a solver for commutative groups to find the proof on its own, and thereby, there will be no extra lemmas to prove afterwards.
This is a something doable because algebraic structures like monoid, group, commutative group and ring have all a very nice property : every element in them has a canonical represent. Thus these kind of proofs can be automatically produced by a tactic by simply computing the normal forms for each side of the equality, and then comparing them with the syntactical equality (Leibniz equality).

\subsection{Our contributions}

@@

In this paper, we focus on some specific theories, which are the following algebraic structures : magmas, semi-group, monoid, group and Abelian group.
The user will just have to prove that his structure satisfies the properties of an algebraic structure, and he will be allowed to invoke the corresponding tactic.
